\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2009-10-02 honors_thesis/thesis/paper.tex
.
First draft of my honors thesis is complete

\documentclass[onecolumn, 12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{amsmath}

\DeclareMathOperator{\URand}{U}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\gbest}{\ensuremath{\Vec{g}}}
\providecommand{\pbest}{\ensuremath{\Vec{p}}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\newcommand{\bib}[1]{../../../bib/#1}

\topmargin .5in

\title{Improving Particle Swarm Optimization \\ in a Parallel World}
\author{Matthew Gardner}
\date{\today}

@@

@@

Optimization problems are ubiquitous in our modern world.  Businesses such as
Google need to decide where to place ads, scientists need to fit models to
data, and airlines need to schedule their flights.  All of those problems are
optimization problems, and optimization techniques can be used to find
solutions.  Optimization algorithms have been derived from all sorts of natural
phenomena.  Some methods look at the landscape of the function as they are
searching and follow hills to the best values, as in the gradient ascent
method.  Other methods take ideas from biology or metallurgy, like genetic
algorithms and simulated annealing.  Particle swarm optimization is a recently
developed optimization technique that draws on ideas from the sociology of
flocking birds.

All of these methods, however, are fundamentally sequential in nature---they
need to be run in a specific order, and because of that, they are typically
only run on one machine.  The problems people want to solve are getting bigger,
and larger problems need to run on multiple machines in parallel.  Parallel
implementations of these algorithms have been attempted, but they do not often
perform as well as sequential versions for the amount of computation performed.
This work focuses on one of those algorithms, particle swarm optimization, and
improving its performance in a large-scale, parallel environment.  The work I
describe in Section~\ref{sec:topology} was done in collaboration with other
students I work with.  The work in Section~\ref{sec:specex} I did by myself.

The rest of this paper is outlined as follows.  In Section~\ref{sec:pso}, I
formally describe the particle swarm optimization algorithm, after giving a
brief description of optimization in general.  In Section~\ref{sec:topology},
I describe the sociological aspect of particle swarm optimization and
improvements that we have made for large swarms.  Section~\ref{sec:specex} then
introduces the idea of speculative execution in general and shows how it can be
done in particle swarm optimization.  Section~\ref{sec:results} shows results
of experiments run, and in Section~\ref{sec:conclusion} I draw some
conclusions.

@@

\subsection{Optimization in General}

The goal of optimization is to find the parameters to a function that either
maximize or minimize the function value.  To take an example from above, Google
needs to find the placements of website advertisements that generate the most
clicks for those websites.  In this example, the placements of advertisements
are the parameters to the function, and the number of clicks generated is the
function to be optimized.  A more mathematical example can be seen in
Figure~\ref{fig:cosine}.  The function shown has a point $x$ for which $f(x)$
is the highest.  Finding the precise location of that point is not a trivial
task, especially for more complicated functions, and it is that problem which
optimization techniques try to solve.

\begin{figure}
  \centering
  \includegraphics{cosine.eps}
  \caption{A complicated function: $\cos(x)\sin(10x)e^{-|x|}$}
  \label{fig:cosine}
\end{figure}

\subsection{Particle Swarm Optimization Specifically}

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart.  It tries to intelligently search a multi-dimensional space by
mimicking the swarming and flocking behavior of birds and other
animals~\cite{kennedy-icnn95}. It is a sociological algorithm that depends on
interaction between particles to quickly and consistently find the optimal
solution to a problem.  The algorithm keeps track of a number of potential
solutions, called particles, which move somewhat randomly through the search
space at each iteration.  When the algorithm is done, the best position that
any of the particles ever found is returned.  Particles remember the best place
they have been, or solution they have evaluated, and when they move they are
attracted back to that place, as well as to the best solution other particles
have seen.  Specifically, the formulas for updating the position $\ppos_t$ and
velocity $\pvel_t$ of a particle at iteration $t$ are as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\gbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called pbest) is the best position the current
particle has seen, and $\gbest$ (called gbest) is the best position any of the
other particles have seen.  \( \coeff_1 \), \( \coeff_2 \), and \(
\constriction \) are parameters with prescribed values required to ensure
convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}~\cite{poli-aea08}.

The sociology in this algorithm defines which other particles form the
``neighborhood'' of a given particle---the other particles whose best solution
it sees (again, that is $\gbest$ in the equations above).  If particle 1 is a
neighbor of particle 2, then particle 1 will be influenced by good values that
particle 2 sees.  But if particle 3 is not a neighbor of particle 1, particle 1
will be completely unaware of good values that particle 3 finds.

There are many ways to define a particle's neighborhood, varying from the
entire rest of the swarm to just one other particle.

The way a neighborhood is defined can have drastic effects on the performance
of the algorithm; the more neighbors each particle has, the faster information
spreads throughout the swarm, and the quicker the algorithm converges.  For
some problems it is possible that the algorithm converges too quickly and gets
caught in a local optimum; it stops on a hill when there is a mountain next to
it.  For those kinds of problems, less communication, or smaller neighborhoods,
is often preferable.

\section{Topology in PSO}
\label{sec:topology}

Much work has been done on the sociological aspect of particle swarm
optimization, trying to determine the best neighborhood structure to use for
various classes of problems.  That work focused on swarms of tens of particles
to several hundred particles on a single machine.  Conclusions were reached
about which neighborhood structures worked best, but only in this limited
context~\cite{bratton-sis07}.  With the advent of parallel processing and the
potential to have swarms of thousands or hundreds of thousands of particles,
some of the earlier work needs rethinking.  This section focuses on
neighborhood structures in very large swarms, especially when the algorithm is
run in parallel on multiple machines.

\subsection{Defining Topology}

Many different neighborhood structures have been proposed and
tested~\cite{kennedy-cec02}.  In the literature, neighborhood structures are
referred to as topologies.  Two of them are more common---the complete or gbest
topology, where all particles communicate with all other particles, and the
ring or lbest topology, where each particle only communicates with its two
immediate neighbors~\cite{bratton-sis07}.

To formalize the description of topologies, we can consider the swarm as a
graph $\{V,E\}$, where the set of particles forms the nodes $V$ on the graph
and the (directed) edges $E$ between nodes represent communication between
particles.  If (1,2) is an edge in the graph, then particle 1 influences
particle 2's gbest, or particle 1 is in particle 2's neighborhood.  The
neighborhood $N_i$ of particle $i$ can be written formally as 
\[N_i = \{p_j|(j,i)\in E\}\]

Neighborhoods can also be thought of in terms of which other particles a
particular particle sends its information to, instead of which ones it gets
information from.  This conceptualization makes the communication a little bit
more explicit (and more directly relates to how to actually run the algorithm)
and will be used to describe topologies throughout the rest of this paper.
$N_i$ then becomes
\[N_i = \{p_j|(i,j)\in E\}\]

The two topologies mentioned previously, complete and ring, would then have the
following neighborhoods for each particle $i$ in a swarm of $n$ particles:
\[\mathrm{Complete:}\ N_i = \{1,2,3,\dots,n\}\]
\[\mathrm{Ring:}\ N_i = \{i-1,i,i+1\}\]

For the benefit of more visual readers, the two topologies are shown in graph
form in Figure~\ref{fig:topgraphs}; notice how many more lines are in the 
complete topology than in the ring topology.

\begin{figure}
  \centering
  \includegraphics{gbest_and_lbest.eps}
  \caption{The complete and ring topologies in graph form.  Figure taken from
  tracer.uc3m.es/tws/pso/neighborhood.html.}
  \label{fig:topgraphs}
\end{figure}

\subsection{Which Topology to Use}

There is a theorem about optimization called ``No Free Lunch''
\cite{wolpert-tec97}. In essence it says that no optimization algorithm can do
better than any other optimization algorithm across all functions.  If you do
better on some functions with a given algorithm that is because you do worse on
other functions.  This applies directly to the question of which topology in
PSO is the ``best,'' because it proves that there really is no ``best''
topology across all functions.  When selecting a topology you must have a
specific type of function in mind, and pick the topology that works best for
that type of function.

An interesting question along those lines is, ``Are the problems that most
people are interested in solving inherently similar?''  If that is the case,
one could design an algorithm that performs better on the problems people care
about, taking a performance hit only on functions that don't matter anyway.
But, because of No Free Lunch, any argument claiming that a particular
algorithm or topology is better than another must make that claim only for a
specific set of functions.

When deciding which topology to use, then, making a wise decision requires
having a certain kind of function in mind.

The complete topology is often preferable for many problems, because it allows
information to be shared quickly throughout the swarm and helps the algorithm
converge more quickly.  However, too much shared information can cause
premature convergence, as mentioned previously, so some researchers are wary of
the complete topology.  In the absence of knowledge about the function to be
optimized, researchers recommended, the ring topology should be used.  It might
take longer, but at least it won't get stuck, they say \cite{bratton-sis07}.

\begin{figure}
  \centering
  \includegraphics{iters_rast.eps}
  \caption{PSO performance on the function Rastrigin with varying numbers of
  particles.}
  \label{fig:rast}
\end{figure}

For a swarm of 50 particles it is accurate that complete often doesn't perform
well, and most researchers were satisfied to stop with a swarm that size.  We
have shown, however, that increasing the number of particles in a swarm will
help the algorithm avoid premature convergence, making complete the best
topology even for somewhat complicated functions~\cite{mcnabb-cec09}.
Figure~\ref{fig:rast} shows the performance of PSO on a commonly used benchmark
function, the Rastrigin function, with varying numbers of particles.  Notice
that with 50 particles the algorithm does get stuck pretty early, not finding
significantly better values after 250 iterations.  This is precisely what many
researchers in the field were saying.  With more particles, however, the
algorithm finds much better values and continues to find better values well
past 250 iterations.

\subsection{Communication Overhead}

In a parallel system, another problem with the complete topology is that
communication between processors is expensive relative to the amount of time it
takes to evaluate the functions many people work with.  A quick look at
Figure~\ref{fig:topgraphs} shows the incredibly large number of messages that
need to be sent between particles for a relatively small swarm.  Communicating
all of the information from every particle at every iteration can take orders
of magnitude more time than doing a simple evaluation and render
parallelization of the algorithm worthless.

We developed a modification to the algorithm along with a new topology that
approximates the complete topology with much less communication.  If each
particle at each iteration randomly chooses a small set of particles to send
its information to, eventually information will be communicated throughout the
whole swarm.  We call this the Random topology, and it is formally defined as
follows:
\[\mathrm{Random:}\ N_i = \{i,U_1,U_2,\dots,U_k\}\]
where $U_j$ is a random number from 1 to $n-1$, and $k$ is the number of 
neighbors to send information to at each iteration.

In the standard particle swarm optimization algorithm, the best neighbor that a
particle is attracted to (the \gbest in the formula above) is taken from its
current neighborhood.  With the random topology we proposed, particles can get
confused, being attracted to a different neighbor at every iteration, and
performance significantly decreases.  We modified the algorithm slightly such
that each particle remembers the best neighbor it has ever had and is attracted
to that position instead of its current best neighbor.  Using this modification
and a random topology, we saw performance that was indistinguishable from
complete with only 5\% of the communication, and the speed up is significant
even in a serial implementation~\cite{mcnabb-cec09}.

Figure~\ref{fig:rastrand} shows the performance of Random versus Complete on
the function Rastrigin for various swarm sizes.  The x-axis shows the number of
particles in the swarm, and the y-axis shows the best function value found
after 500 iterations.  It can be seen from the graph that Random does a very
good job at approximating the Complete topology while required only a small
fraction of the communication.

\begin{figure}
  \centering
  \includegraphics{randstar.eps}
  \caption{Random with various amounts of communication versus the Complete
  topology on the function Rastrigin.}
  \label{fig:rastrand}
\end{figure}

\subsection{Subswarms}

We have run all of the experiments I've talked about so far with swarms of up
to 4000 particles.  All of them were run with a serial implementation of
particle swarm optimization on a single machine, though the results can be used
in designing parallel algorithms.  With more than 4000 particles the algorithm
becomes very computationally expensive on a single machine.  We have
implemented a parallel version of particle swarm optimization using Google's
MapReduce framework for parallel computation~\cite{dean-osdi04}.  Soon we will
have it working well enough to run swarms of hundreds of thousands of particles
on hundreds of machines at a time, and we will further explore the issues
involved with parallel particle swarm optimization.

Many new possibilities for topologies arise when the algorithm is run in
parallel with large swarms.  We have proposed, but not yet fully tested, a
topology of subswarms, where each of a hundred machines would evaluate 500 or
1000 particles and send only its best value to all other machines, drastically
reducing the amount of inter-processor communication.  In our formal notation,
the subswarms topology can be described as

\[\mathrm{Subswarms:}\ N_i = \{j|j\%k = i\%k\}\]
where $k$ is the number of subswarms in the swarm, and \% is the mod operator.
For a small example, in a swarm of 12 particles with 3 subswarms, the sets of
neighbors would be \{1, 4, 7, 10\}, \{2, 5, 8, 11\}, and \{3, 6, 9, 12\}.

Our preliminary results show that this approach is very
promising~\cite{mcnabb-cec09}.  Figure~\ref{fig:subswarms} shows the results
of using the Subswarms topology with 400 particles in each swarm, versus just
using 400 particles in a Complete topology.  Subswarms clearly outperforms
Complete in this case.

\begin{figure}
  \centering
  \includegraphics{islands_iters_rast.eps}
  \caption{Subswarms versus Complete topologies on the function Rastrigin.}
  \label{fig:subswarms}
\end{figure}

\section{A Different Speed up---Speculative Execution}
\label{sec:specex}

\subsection{Speculative Exectuion in General}

Speculative execution in a program is the execution of code that may or may not
end up actually being needed.  Modern processors routinely do this when a
conditional branch is encountered---they try to predict which branch will
actually be needed and continue their execution.  If it turns out that the
prediction was incorrect, the work is discarded.  But if the prediction was
right, the program can execute much faster than if it had waited on the branch.  
\subsection{Speculative Execution in Particle Swarm Optimization}

In many optimization algorithms speculative execution is impossible, because
the next iteration of the algorithm depends on the evaluation of the function
during the previous iteration.  In simulated annealing, for instance, the
probability of accepting a new position depends on the difference in function
value between the new position and the current one.  In gradient descent, the
gradient of the function must be determined (which includes actually evaluating
it) to know how far to move before sampling the function again.

Particle swarm optimization is rather unique in that it does not require the
evaluation of the function to know what the next position of each particle will
be.  It simply depends on \emph{which} particle ended up having the best
position and whether or not the current particle found a better spot than it
had seen before.  That means you can just compute all possible next
positions---all combinations of which particle is the new gbest and whether or
not pbest was updated (\gbest\ and \pbest\ from equation
\eqref{eq:velupdate})---and evaluate each position in parallel.  That computes
two iterations at the same time.

\subsection{Is is worth it?}

It uses a lot of extra work to get the second iteration, but for some functions
performing more iterations makes more progress than adding extra particles.

\section{Conclusion}
\label{sec:conclusion}

Thank you for reviewing my proposal.  I look forward to exploring speculative
execution in particle swarm optimization.

>>>>

\documentclass[conference,letterpaper]{IEEEtran}

\usepackage{cite}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}

\IEEEoverridecommandlockouts

% AML bibliography references
\newcommand{\bib}[1]{/aml/home/mjg82/bib/#1}

% Better references, I think
\renewcommand{\sec}[1]{Section~\ref{sec:#1}}
\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
\newcommand{\alg}[1]{Algorithm~\ref{alg:#1}}

% Algorithmic changes
\renewcommand{\algorithmicforall}{\textbf{for each}}

%% PSO Stuff
\DeclareMathOperator{\URand}{U}
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}

%SpecExPSO Stuff
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}

\title{\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization%
\thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department
of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602
(phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}%
}

\date{}

\author{Matthew Gardner, Andrew McNabb, and Kevin Seppi}

@@

\begin{abstract}

Particle swarm optimization (PSO) has previously been parallelized only by 
adding more particles to the swarm or by parallelizing the evaluation of the
objective function.  However, some functions are better optimized with more
iterations and fewer particles.  Accordingly, we take inspiration from 
speculative execution commonly performed in modern processors and propose what
we call speculative evaluation in PSO.  Future positions of the particles are
speculated and evaluated in parallel with current positions, performing two
iterations of PSO at once.

We also propose another way of making use of these speculative particles,
keeping the best position found instead of the position that PSO actually would
have taken.  We show results that for a number of functions, speculative
evaluation can have dramatic improvements over adding additional particles to
the swarm.

\end{abstract}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  Even in desktop machines, systems with 4 or 8 cores are common
place.  In order to efficiently utilize these resources for computationally
intensive problems, algorithms need to run in parallel.

Within the last few years, researchers have begun to recognize the need to
develop parallel implementations of PSO, publishing many papers on the subject.
The methods they have used include various synchronous algorithms
\cite{belal-ijicis04,chu-sci06,jin-aps05,parsopoulos-aia04,schutte-ijnme04},
asynchronous algorithms \cite{koh-ijnme06,mostaghim-report06,venter-wcsmo05},
and formulating PSO in Google's MapReduce framework \cite{mcnabb-cec07}.
Parallelizing the evaluation of the objective function can also be done in some
cases, though that is not an adaption of the PSO algorithm and thus is not the
focus of this paper.

All of these parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new subswarms.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function--additional processors add more particles, not
more iterations.  However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait, guess which way the branch will go and start
executing.  If the processor guessed right, execution is much farther ahead
than if it had just idly waited on the branch.  If it guessed wrong, execution
is right where it would have been anyway.  This is called speculative
execution or branch prediction.

The idea of speculative execution can be applied to some types of optimization
algorithms.  When the next sampling position of the algorithm does not depend
explicitly on the value of the objective function at the current position,
future positions can be speculated and evaluated in parallel with the current
position.  This could not be done in gradient methods, for instance, where the
step size for sampling depends on the gradient at the current position, but it
has been done in simulated annealing~\cite{witte-tpds91}.

Because the motion of particles in PSO depends only on the \emph{positions} of
the particle and its neighbors, not the \emph{value} of the objective function
at those positions, speculative evaluation is possible.  The relative values of
the positions of the particle's neighbors do matter, in that the best one is
used as the neighborhood best position, but it's which position has the highest
value, not what that value is, that matters.  We then propose a new method of
parallelizing PSO in which additional processors are used to perform more
iterations, not add particles to the swarm.

Because we do not have to wait for a particle's position to be evaluated to know
where it might move to next, we can compute the set of all possible next
positions and send them to be evaluated along with the particle's current
position.  Once the current position of each particle has been evaluated, we
can determine which of the speculatively evaluated positions was correct.  We
have thus evaluated two iterations at once.  This can easily get unwieldy if
the set of next possible positions for each particle is large, but a wise
choice of topology limits that set to a reasonable size.

The original PSO can be reproduced exactly, two iterations at a time, using
this technique.  However, in strictly using speculative evaluation to reproduce
the original PSO, a lot of positions are speculatively evaluated whose values
get ignored after deciding which branch was the correct one.  A better use of
those speculative evaluations may be to incorporate the information into the
swarm, no longer exactly reproducing PSO but perhaps producing better results.

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once.  In \sec{extra}, we discuss a variant of the
speculative algorithm which uses all of the speculative evaluations performed
instead of throwing away all but one for each particle.  We then show in
\sec{results} a comparison between the original PSO algorithm, our speculative
algorithm, and its variant.

@@

Particle swarm optimization was proposed in 1995 by James Kennedy and Russell
Eberhart~\cite{kennedy-icnn95}.  The algorithm is used to intelligently search
a multi-dimensional space by mimicking the swarming and flocking behavior of
birds and other animals. It is a social algorithm that depends on interaction
between particles to quickly and consistently find the optimal solution to some
objective function.

The motion of particles through the search space has three components: an
inertial component that gives particles momentum as they move, a cognitive
component where particles remember the best solution they have found and are
attracted back to that place, and a social component by which particles are
attracted to the best solution that any of their neighbors have found.

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\nbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
\coeff_2 \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  We use the Ring topology, where each particle
is neighbors only with two other particles, and a parallel adaptation of the
Complete topology~\cite{mcnabb-cec09}, where each particle is neighbors with
all other particles.  These are the two most commonly used topologies in the
literature and have been shown to work well in practice~\cite{bratton-sis07}.

\section{Speculative Evaluation in PSO}
\label{sec:sepso}
\subsection{Overview}

The PSO algorithm just described can be parallelized by having each additional
processor add a particle or a set of particles to the swarm.  But for some
functions, a more desirable use of resources is to speed up the algorithm,
performing more iterations instead of adding more particles.  Barring
parallelizing the evaluation of the objective function itself, which may not be
possible, previous techniques have not addressed this issue.

In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.  If we can guess
what a particle's next position will be from its current position, we can
evaluate both in parallel and perform two iterations of PSO at once.  Lacking
a good way to guess a particle's next position, we can at least enumerate all
possible next positions of the particle, evaluate all of them at once, and one
of them will be right.

While in some algorithms enumerating all possible next sampling positions would
be impossible or extremely difficult, in PSO all we need to know are the
current positions of the particle and its neighbors, from which we can easily
enumerate and evaluate all possible positions of that particle at the next
iteration.  The defining components of a particle's motion
(see~\eqref{eq:velupdate}) are $\pvel_t$, $\ppos_t$, $\pbest$, and $\nbest$.
For a particle at iteration $t-1$, $\pvel_t$ and $\ppos_t$ are defined by
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}.  The particle already
knows its personal best $\pbest$ at iteration $t-1$, but it could be updated to
$\ppos_t$ if its position at iteration $t$ has a better value than its current
personal best.  Likewise, the particle knows its neighborhood best $\nbest$ at
iteration $t-1$, though it could be updated by any of its neighbors' positions
at iteration $t$.

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has (a particle either updates its personal best or it
does not, and one of its $n$ neighbors must update its neighborhood best, or
there is no neighborhood best update--$n+1$ possibilities).  If all of those
speculative positions are evaluated in parallel along with the original,
non-speculative position of the particle at time $t-1$, then one of the
speculative positions has to be the correct one, and two iterations have been
evaluated at once.

Because $2(n+1)$ speculative evaluations must be performed for each particle,
the choice of topology is an important one.  The use of the Complete topology,
where every particle is a neighbor to every other particle, would require on
the order of $p^2$ speculative evaluations per iteration, where $p$ is the
swarm size.  Clearly it is much more desirable to have a sparse topology, where
$O(np)$ is much smaller than $O(p^2)$.  However, some functions are better
optimized with the Complete topology and the quick spread of information it
entails than with sparse topologies.  Accordingly, we use the Random topology
described in~\cite{mcnabb-cec09}, which has been shown to effectively simulate
the Complete topology.  In that topology, 2 neighbors are chosen at random from
the swarm for each particle at each iteration.  Because a particle is also
always its own neighbor, $n$ is $3$ in that case.  In~\sec{results} we report
the results for using both the Ring topology (where $n$ again is $3$) and the
Random topology on a number of common benchmark functions.

It is not trivial in some parallel architectures to determine which speculative
position was the correct next position of each particle.  First we discuss the
relatively easy case of a centralized parallel PSO algorithm.  In such an
architecture, the master keeps track of all necessary information with only
trivial message passing needed.  Then we discuss the more complicated case of a
distributed algorithm, where each particle is on its own and needs to send and
receive messages to and from other particles.  Finally we discuss the further
complications of a dynamic topology, where a particle's neighbors change from
one iteration to another.

\subsection{Terminology}

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
$\sset_{t+1}$.  $\n_t$ is a neighbor of particle $\p_t$.  A set of particles is
$\pset$, whereas a single particle is simply $\p$.

We separate each iteration of PSO into several steps.  First there is the
motion step, where a particle updates its position and velocity.  Then a
particle's position is evaluated, and the particle updates its current value
and its personal best.  Finally, a particle gets information from its neighbors
and updates its neighborhood best.

A particle at iteration $t-1$ that has been moved to iteration $t$, but whose
position has not yet been evaluated, is denoted as $\noeval{\p}_t$.  Once its
position has been evaluated, but it has still not yet received information from
its neighbors, it is denoted as $\nonbest{\p}_t$.  Only when the particle has
updated its neighborhood best is it a complete particle at iteration $t$.  It is
then simply denoted as $\p_t$.

\subsection{Centralized Algorithms}

In a centralized, or Master-Slave, parallel PSO algorithm, one machine, the
master, keeps track of all necessary information, and all other machines are
merely used to evaluate the objective function at various positions as directed
by the master~\cite{belal-ijicis04}.  To perform speculative evaluation in such
a scheme, the master generates the positions to evaluate speculatively and
decides which one to accept for each particle.

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with.  Each of those
combinations defines a speculative particle, $\s_t$, that can be moved using
the motion equations to obtain $\noeval{\s}_{t+1}$.  The result is a set of
particles $\noeval{\pset}_t$, and for each particle a set of speculative
children $\noeval{\sset}_{t+1}$, which can all be evaluated at once.

The master then has the slaves evaluate the particles.  Once all particles,
speculative and original, have been evaluated and the values reported to the
master, the master sorts out which speculative child of each particle was the
correct one.

This is done by first updating each particle $\noeval{\p}_t$ to
$\nonbest{\p}_t$ by associating the value determined by the slave to the
particle it came from, and updating that particle's personal best if the value
reported is better than its previous personal best value.  Then each
$\nonbest{\p}_t$ must be updated to $\p_t$ by updating its neighborhood best
position and value with information about each of its neighbors.  This is
simply the original PSO algorithm.

Then, with the set $\pset_t$, the branch that was actually taken by each
particle can be determined, and the correct speculative child can be chosen.
The branches depended entirely on whether or not the personal best of $p_t$ was
updated and which particle, if any, updated the particle's neighborhood best.
Both of those items are now known, and the child that matches the branch is
selected.

The parent $\p_t$ must pass its personal best value to the child, as the child
knows only the position that it guessed, not the function value at that
position.  It is possible that both $\p_t$ and $\s_{t+1}$ updated their
personal bests, but $\p_t$'s value was better.  For example, suppose that
$\p_{t-1}$ has a personal best value of 3.  $\noeval{\p}_t$ was created, and
$\noeval{\s}_{t+1}$ was moved assuming that $\p_t$ had updated its personal
best with its position at time $t$.  Then both $\noeval{\p}_t$ and
$\noeval{\s}_{t+1}$ were evaluated, with values 1 and 2, respectively.
$\nonbest{\s}_{t+1}$ would think that its current position was its personal
best, as the value it found--2--was better than its previous personal best
value of 3.  It needs to receive the personal best value from its parent to
know that its personal best position $\pbest$ is actually the position of
$\p_t$, not $\s_{t+1}$.

The parent also needs to pass the value of the neighborhood best that the child
guessed.  The child only knows the position and needs the value in order to
make future comparisons between neighborhood best positions.

Upon picking the correct branch for each particle and updating the children's
personal best and neighborhood best value (from iteration $t$), the result is
the set $\nonbest{\pset}_{t+1}$, as the particles are now no longer
speculative.  What remains is to update the neighborhood best of those
particles from their neighbors (from iteration $t+1$), as above, to obtain
$\pset_{t+1}$.  That set of particles can subsequently be used to produce the
sets $\noeval{\pset}_{t+2}$ and $\noeval{\sset}_{t+3}$, and the process repeats
itself.

The algorithm is outlined in \alg{centralized}.

\begin{algorithm}
  \caption{Speculative Evaluation in a Centralized PSO}
  \label{alg:centralized}
  \begin{algorithmic}[1]
	\STATE Move all $\p_{t-1}$ to $\noeval{\p}_t$
	\STATE For each $\noeval{\p}_t$, get its neighbors $\noeval{\nset}_t$ and
	  generate $\noeval{\sset}_{t+1}$
	\STATE Evaluate all $\noeval{\p}_t$ and $\noeval{\sset}_{t+1}$ in parallel
	\STATE Update personal best for each $\noeval{\p}_t$ and
	  $\noeval{\s}_{t+1}$, creating $\nonbest{\p}_t$ and $\nonbest{\s}_{t+1}$
	\STATE Update neighborhood best for each $\nonbest{\p}_t$, creating
	  $\pset_t$
	\FORALL{$\p_t$}
	\STATE Pick $\nonbest{\s}_{t+1}$ from $\nonbest{\sset}_{t+1}$ that matches
	  the branch taken by $\p_t$
	\STATE Pass along personal and neighborhood best values obtained by $\p_t$,
	  making $\nonbest{\p}_{t+1}$
	\ENDFOR
	\STATE Update neighborhood best for each $\nonbest{\p}_{t+1}$, creating
	  $\pset_{t+1}$
	\STATE Repeat
  \end{algorithmic}
\end{algorithm}

\subsection{Distributed Algorithms}

In a distributed parallel PSO algorithm, individual processors not only perform
evaluations of particles, but also their movement.  The information for each
particle is not held by a central machine that directs the algorithm; instead,
each processor has the information for the particle or particles that it is in
charge of and must perform the steps of the algorithm for the particles it
manages.  Messages such as values and positions for the neighborhood best are
sent between processors.  There may be (in fact there probably must be) some
machine that collects information from all of the particles and outputs the
result of the algorithm, though that machine's importance is much less than in
centralized algorithms.

To perform speculative evaluation in a distributed PSO algorithm, there must be
some way to have processors evaluate the speculative children of particles,
without giving the speculative particles the same treatment as actual
particles.  Perhaps that can be done by assigning each particle a set of
machines instead of a single machine, and the particle directs its extra
machines to evaluate its speculative children.  In the MapReduce framework,
each particle and speculative child has its own Map task, during which the
particle is evaluated, with the speculative children sending their output back
to the original particle's Reduce task and any neighbors it has.  The Reduce
task for each particle gets all of the messages from its neighbors, its
speculative children, and the speculative neighbors of its speculative
children, and produces the next particle and children for another Map task.

No matter which framework is used the message passing should be similar, and
that is what we describe here.

A processor that is controlling a single particle $\p_{t-1}$ must first move
the particle to $\noeval{\p}_t$ and produce the particle's speculative children
$\noeval{\sset}_{t+1}$.  This is done in the same way as described above.
However, in order to produce $\noeval{\sset}_{t+1}$, the processor needs
information about the particle's neighbors, so there must be some message
passing to get that information.  Particularly, the information that the
processor needs is the position of each of the particle's neighbors at
iteration $t$.

This can be accomplished in two ways, either by performing several rounds of
very simple message passing, or by performing a single round with a lot more
messages.
